One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-08 paper/discussion.tex
update

\subsection{Role of self-gravity and disc structure} 
We note that the background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity, and we only considered a self-gravitating perturbation to
obtain an appropriate wavenumber $k$ needed to evaluate
Eq. \ref{theoretical_rate}. Thus, self-gravity is not directly
responsible for instability. However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, the main role of disc structure and self-gravity is to
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

@@

\subsubsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. To estimate this effect, we port basic results from
disc-planet theory \citep[see, e.g.][and references
therein]{papaloizou07}.

Let us regard the one-arm spiral disturbance located in the inner disc as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We expect
it to exert a torque on the exterior disc at Lindblad  
and co-rotation resonances. At the outer Lindblad resonance (OLR), this
torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
\left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
\end{align}   
where a Keperian disc has been assumed and subscript $L$ denotes
evaluation at $R=R_L$. (The inner Lindblad resonance does not exist
for the pattern speeds observed in our simulations.)

@@

For the co-rotation torque, we use the result
\begin{align}
  \Gamma_C = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{C}
\end{align}
for a Keplerian disc, where subscript $C$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
  \Gamma_C = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
    \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
      - \bar{R}\right)^2}.   
\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \frac{\gamma_C}{\Omega_p} \sim 0.010\Omega_p,
\end{align}
which is again small compared to the measured growth rates.

>>>>

\subsection{The role of self-gravity and disc structure} 
The background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity explicitly. In the above discussion, we only considered a self-gravitating perturbation to 
obtain an appropriate wavenumber $k$ needed to evaluate
Eq. \ref{theoretical_rate}. Self-gravity is not directly
responsible for instability. However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

In order to confine an $m=1$ spiral between two $Q$-barriers, we
should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
Keperlian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
\begin{align}
  \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
\end{align}
Then it may be possible to have two $Q$-barriers when the $Q$ profile
of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
for decreasing (increasing) $R$. One possibility is a gap
opened by a giant planet in a constant-$Q$ disc. In that case $Q$
rises rapidly towards the gap centre since it is a region of low
surface density.  %multi spirals associated with one planet

@@

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. To estimate this effect, we port basic results from
disc-planet theory \citep[see, e.g.][and references
therein]{papaloizou07}.

Let us regard the one-arm spiral disturbance located in the inner disc as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We expect
it to exert a torque on the exterior disc at the Lindblad  
and co-rotation resonances. At the outer Lindblad resonance (OLR), this
torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
\left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
\end{align}   
where a Keperian disc has been assumed and subscript $L$ denotes
evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
exist for the pattern speeds observed in our simulations.)

@@

For the co-rotation torque, we use the result
\begin{align}
  \Gamma_C = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{C}
\end{align}
for a Keplerian disc, where subscript $C$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
    \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
      - \bar{R}\right)^2}.   
\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_C\sim 0.010\Omega_p,
\end{align}
which is again small compared to the measured growth rates.

%not important for our models, but may be important if outer disc is
%self-gravitating too

### 2014-11-08 paper/paper.tex
update

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\newcommand{\varA}[1]{{\operatorname{#1}}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we speculate it may also
  affect low-frequency one-arm spirals in non-self-gravitating
  discs.

  % or confirm in 3D
  %, taken as a toy model of a 
  %protoplanetary disc containing a finite-size massive dead zone.
   % Theoretical models of circumstellar or protoplanetary discs indicate
  % the possible existence of `dead zones' --- regions in the disc
  % where mass accretion is greatly reduced because of inefficient
  % magneto-hydrodynamic turbulence that would otherwise enable 
  % accretion. It has been suggested that mass built-up in a dead zone
  % may lead to gravitational instability. However, what type of
  % gravitational instability is relevant in this context has not been
  % clarified. 
  %
  % Our simulations demonstrate the 
  % importance of global gravitational interaction between a 
  % self-gravitating dead zone and the exterior, non-self-gravitating
  % parts of the disc. This non-local effect, due to disc structure, 
  % exists even when the total disc mass is below the threshold for
  % global effects found in
  % previous studies. 
\end{abstract}

\begin{keywords}
  planetary systems: formation --- planetary systems:
  protoplanetary discs
\end{keywords}

\section{Introduction}\label{intro}

\input{model}
\input{results}
\input{results_3d}
\input{discussion}
%\input{summary}

@@

\appendix
\input{appendix}
%\input{appendix2}

>>>>

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\pasj}{PASJ}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\newcommand{\varA}[1]{{\operatorname{#1}}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we speculate it may also
  affect low-frequency one-arm spirals in non-self-gravitating
  discs. 
\end{abstract}

\begin{keywords}
  accretion, accretion discs, hydrodynamics, instabilities, methods: numerical, protoplanetary discs 
\end{keywords}

\section{Introduction}\label{intro}
% Protoplanetary discs are expected to contain complex structures
% \citep{armitage11}, which directly impact planet formation theories.  
One of the exciting developments in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{muto12,grady13,christiaens14,avenhaus14,boccaletti14}.

% planet is popular answer for spiral structure
An attractive explanation for asymmetries in circumstellar discs is to
invoke disc-planet interaction. In particular, spiral structures
naturally arise from the gravitational interaction between a
planet and the gaseous protoplanetary disc it is embedded in
\citep[see, e.g.][for a recent review]{baruteau13b}. Thus, the
presence of spiral arms in circumstellar discs could be signposts of 
planet formation.

% but also natural in GI. when/where does it happen? 
Spiral arms are also naturally associated with gravitational
instability in differentially rotating discs.

%original motivation for this work (DZGI)

%specifically m=1 (history, conditions)

\input{model}
\input{results}
\input{results_3d}
\input{discussion}
\input{summary}

@@

\appendix
\input{appendix}
